A tutorial on Generalised Additive Mixed Effects Models for bilingualism research

### 2024-10-10 manuscript/manuscript.qmd
update affiliations
revise manuscript

Joseph V. Casillas (Rutgers University, USA)

@@

Research on language acquisition first documented over-regularization errors in the 60's [@ervin_miller1963; @cazden1968acquisition]. Subsequent research found that over-regularization errors tended to manifest in a developmental curve, and is now referred to as the *U-shaped developmental curve* [@williams2022u]. In most cases, U-shaped development is considered to be a three-step process, which begins with accurate performance, followed by a period in which performance dips, and then subsequently becomes accurate again [@carlucci2013necessity]. Interestingly, U-shaped curves are observed in numerous cognitive-developmental and learning contexts, particularly with regard to child development, e.g., understanding temperature, weight conservation, object permanence, and facial recognition [See @carlucci2013necessity], which, in turn, has fostered fruitful research and debate in Cognitive Science regarding learning models [e.g., @marcus1992overregularization; @rumelhart2014learning, among others].

@@

In the following example, we consider the production of Spanish vowels in the aforementioned groups of bilingual participants. Interestingly, to our knowledge, all of the current research investigating the cross-linguistic influence of vowel reduction processes in learners of Spanish have utilized modelling strategies that rely upon formant measurements at a single time point, typically the mid-point of the vowel, rather than scrutinizing the entire trajectory of the spectral envelope. In our view, the dynamic nature of vowel formants poses an interesting use-case for modelling using non-linear methods, such as GAMs. To this end, we have simulated phonetic data from three aforementioned groups of bilinguals: simultaneous English/Spanish bilinguals, beginner late English/Spanish bilinguals and advanced late English/Spanish bilinguals. The data contains the Euclidean Distance (EuD) of three Spanish vowels /a, i, u/ from the vowel space centroid, taken from word-final non-stressed vowels, along nine equidistant time points. The measure is a good proxy for vowel reduction phenomena, which is to be expected in late bilinguals.

>>>>

Joseph V. Casillas (Department of Spanish and Portuguese, Rutgers University, USA)

@@

Research on language acquisition first documented over-regularization errors in the 60's [@ervin_miller1963; @cazden1968acquisition]. Subsequent research found that over-regularization errors tended to manifest in a developmental curve, and is now referred to as the *U-shaped developmental curve* [@williams2022u; @Botezatu2024]. In most cases, U-shaped development is considered to be a three-step process, which begins with accurate performance, followed by a period in which performance dips, and then subsequently becomes accurate again [@carlucci2013necessity]. Interestingly, U-shaped curves are observed in numerous cognitive-developmental and learning contexts, particularly with regard to child development, e.g., understanding temperature, weight conservation, object permanence, and facial recognition [See @carlucci2013necessity], which, in turn, has fostered fruitful research and debate in Cognitive Science regarding learning models [e.g., @marcus1992overregularization; @rumelhart2014learning, among others].

@@

In the following example, we consider the production of Spanish vowels in the aforementioned groups of bilingual participants. Interestingly, to our knowledge, all of the current research investigating the cross-linguistic influence of vowel reduction processes in learners of Spanish have utilized modelling strategies that rely upon formant measurements at a single time point, typically the mid-point of the vowel, rather than scrutinizing the entire trajectory of the spectral envelope (See @FuenteIglesias2022 and @Amengual2020). In our view, the dynamic nature of vowel formants poses an interesting use-case for modelling using non-linear methods, such as GAMs. To this end, we have simulated phonetic data from three aforementioned groups of bilinguals: simultaneous English/Spanish bilinguals, beginner late English/Spanish bilinguals and advanced late English/Spanish bilinguals. The data contains the Euclidean Distance (EuD) of three Spanish vowels /a, i, u/ from the vowel space centroid, taken from word-final non-stressed vowels, along nine equidistant time points. The measure is a good proxy for vowel reduction phenomena, which is to be expected in late bilinguals.

### 2024-10-10 manuscript/manuscript.tex
update affiliations
revise manuscript

Research on language acquisition first documented over-regularization
errors in the 60's (Cazden, 1968; Ervin \& Miller, 1963). Subsequent
research found that over-regularization errors tended to manifest in a
developmental curve, and is now referred to as the \emph{U-shaped
developmental curve} (Williams et al., 2022). In most cases, U-shaped
development is considered to be a three-step process, which begins with
accurate performance, followed by a period in which performance dips,
and then subsequently becomes accurate again (Carlucci \& Case, 2013).
Interestingly, U-shaped curves are observed in numerous
cognitive-developmental and learning contexts, particularly with regard
to child development, e.g., understanding temperature, weight
conservation, object permanence, and facial recognition (See Carlucci \&
Case, 2013), which, in turn, has fostered fruitful research and debate
in Cognitive Science regarding learning models (e.g., Marcus et al.,
1992; Rumelhart \& McClelland, 1986, among others).

@@

In the following example, we consider the production of Spanish vowels
in the aforementioned groups of bilingual participants. Interestingly,
to our knowledge, all of the current research investigating the
cross-linguistic influence of vowel reduction processes in learners of
Spanish have utilized modelling strategies that rely upon formant
measurements at a single time point, typically the mid-point of the
vowel, rather than scrutinizing the entire trajectory of the spectral
envelope. In our view, the dynamic nature of vowel formants poses an
interesting use-case for modelling using non-linear methods, such as
GAMs. To this end, we have simulated phonetic data from three
aforementioned groups of bilinguals: simultaneous English/Spanish
bilinguals, beginner late English/Spanish bilinguals and advanced late
English/Spanish bilinguals. The data contains the Euclidean Distance
(EuD) of three Spanish vowels /a, i, u/ from the vowel space centroid,
taken from word-final non-stressed vowels, along nine equidistant time
points. The measure is a good proxy for vowel reduction phenomena, which
is to be expected in late bilinguals.

@@

@@

@@

>>>>

Research on language acquisition first documented over-regularization
errors in the 60's (Cazden, 1968; Ervin \& Miller, 1963). Subsequent
research found that over-regularization errors tended to manifest in a
developmental curve, and is now referred to as the \emph{U-shaped
developmental curve} (Botezatu et al., 2024; Williams et al., 2022). In
most cases, U-shaped development is considered to be a three-step
process, which begins with accurate performance, followed by a period in
which performance dips, and then subsequently becomes accurate again
(Carlucci \& Case, 2013). Interestingly, U-shaped curves are observed in
numerous cognitive-developmental and learning contexts, particularly
with regard to child development, e.g., understanding temperature,
weight conservation, object permanence, and facial recognition (See
Carlucci \& Case, 2013), which, in turn, has fostered fruitful research
and debate in Cognitive Science regarding learning models (e.g., Marcus
et al., 1992; Rumelhart \& McClelland, 1986, among others).

@@

In the following example, we consider the production of Spanish vowels
in the aforementioned groups of bilingual participants. Interestingly,
to our knowledge, all of the current research investigating the
cross-linguistic influence of vowel reduction processes in learners of
Spanish have utilized modelling strategies that rely upon formant
measurements at a single time point, typically the mid-point of the
vowel, rather than scrutinizing the entire trajectory of the spectral
envelope (See de la Fuente Iglesias \& P'erez Castillejo (2022) and
Amengual \& Simonet (2020)). In our view, the dynamic nature of vowel
formants poses an interesting use-case for modelling using non-linear
methods, such as GAMs. To this end, we have simulated phonetic data from
three aforementioned groups of bilinguals: simultaneous English/Spanish
bilinguals, beginner late English/Spanish bilinguals and advanced late
English/Spanish bilinguals. The data contains the Euclidean Distance
(EuD) of three Spanish vowels /a, i, u/ from the vowel space centroid,
taken from word-final non-stressed vowels, along nine equidistant time
points. The measure is a good proxy for vowel reduction phenomena, which
is to be expected in late bilinguals.

@@

\bibitem[\citeproctext]{ref-Amengual2020}
Amengual, M., \& Simonet, M. (2020). Language dominance does not always
predict cross-linguistic interactions in bilingual speech production.
\emph{Linguistic Approaches to Bilingualism}, \emph{10}(6), 847--872.
\url{https://doi.org/10.1075/lab.18042.ame}

@@

\bibitem[\citeproctext]{ref-Botezatu2024}
Botezatu, M. R., Misra, M., \& Kroll, J. F. (2024). Proficiency in a
second language influences processing of print-to-sound mappings.
\emph{Linguistic Approaches to Bilingualism}, \emph{14}(3), 285--309.
\url{https://doi.org/10.1075/lab.21063.bot}

@@

\bibitem[\citeproctext]{ref-FuenteIglesias2022}
de la Fuente Iglesias, M., \& P'erez Castillejo, S. (2022). L1 phonetic
permeability and phonetic path towards a potential merger: The case of
galician mid vowels in bilingual production. \emph{Linguistic Approaches
to Bilingualism}, \emph{12}(2), 191--219.
\url{https://doi.org/10.1075/lab.19094.del}
